Make any stupid plot you want

### 2021-01-19 draft/sections/intro.tex
new pdf
removed minted, fleshed out fiber bundles

\subsection{ What is a viz}?
Acquired codes of meaning

@@

\subsection{Terminology}
\begin{figure}[h!]
%%remake properly at some point
\includegraphics[width=\textwidth]{figures/intro/flowchart.png} 
\caption{To go from a dataset to a visualization, the data is subset based on a set of constraints (the invariant). The resulting subset becomes the components that are visualized, but the choice of visualization is dependent on the type and structure of the component variables.}
\label{fig:flowchart}
\end{figure}

Given a dataset, we need to decide what subset of the data to visualize. Bertin describes the set of constraints used to subset the data as the \textit{invariant}. Formally, the \textit{invariant} is the set of shared characteristics of the data being visualized. When these constraints are applied to the dataset, the resulting subset is what will become the \textit{components} of the visualization \cite{bertin_semiology_2011}. In figure~\ref{fig:iris_scatter}, the \textit{invariant} common to all the data being visualized is "sepal length", "petal length", and "species" and the \textit{components} are the measurements of these variables. As shown in figure~\ref{fig:flowchart}, the final step in creating a visualization is choosing how to encode the
components using retinal (visual) variables.

\subsubsection{Retinal (Visual) Variables}
\begin{figure}[h!]
\includegraphics[width=1\textwidth]{figures/intro/retinal_variables.png}
\caption{Retinal variables are a codification of how position, size, shape, color and texture are used to illustrate variations in the \textit{components} of a visualization. This tabular form of Bertin's retinal variables is from Understanding Graphics \cite{_information_????} who reproduced it from \textit{Making Maps: A Visual Guide to Map Design for GIS} \cite{krygier_making_2005}}
\label{fig:retinal_variables}
\end{figure}

@@

\subsubsection{Data Type and Structure}

\begin{figure}[h!]
 \includegraphics[width=\textwidth]{figures/intro/munzner_datatypes}
\caption{Keys are unique lookup values used to find individual observations in the dataset. Keys are positional references, and can be coordinates on a map or unique values such as a primary key in a database or a (time, latitude, longitude) index in a data cube. Image modified from a diagram from Munzner's website \cite{_visualization_????}}
\label{fig:munzner_datatypes}
\end{figure}

As shown in figure~\ref{fig:flowchart}, there are multiple ways to translate data into pictures. A map is always an option, except when the observations do not have associated coordinates in a physical plane. Tamara Munzner provides a way to distinguish between these datasets using $\{key, value\}$ designations \cite{munzner_visualization_2014}. Munzner defines \textit{values} as measurements of interest in the dataset, analogous to dependent variables in statistics. She defines \textit{keys} as indexes that can be used to look up values, analogous to independent variables in statistics and dimensions in computer science. Figure~\ref{fig:munzner_datatypes} illustrates how these keys are used to look up variables in a dataset: 
\begin{itemize}
	\item map: keys are the coordinates of the points
	\item table: row index, database primary key
	\item data cube: row, column, etc. (.e.g. $i,j,k$) index
\end{itemize}

Expanding on Munzner's key and value semantics, in many datasets the keys are discrete variables like time or geophysical locations sampled from a continuous curve, surface, or field. While these observations are discrete samples from the continuous space, often the continuous (functional) characteristic\cite{ramsay_functional_2006,muller_functional_2006} of the observational space is also of interest. Besides quantitative discrete, quantitative continuous, or categorical measurement type considerations, the choice of visualization is also influenced by the measurement being on an interval, ratio, nominal, or categorical scale.

>>>>

\subsection{ What are the the basic requirements of a visualization library?}
Acquired codes of meaning, dubois

@@

\subsection{Terminology}
\textcolor{blue}{Is currently C\&P'ed from my second exam so needs some messaging, but }
\begin{figure}[h!]
%%remake properly at some point
\includegraphics[width=\textwidth]{figures/intro/flowchart.png} 
\caption{To go from a dataset to a visualization, the data is subset based on a set of constraints (the invariant). The resulting subset becomes the components that are visualized, but the choice of visualization is dependent on the type and structure of the component variables.}
\label{fig:flowchart}
\end{figure}

Given a dataset, we need to decide what subset of the data to visualize. Bertin describes the set of constraints used to subset the data as the \textit{invariant}. Formally, the \textit{invariant} is the set of shared characteristics of the data being visualized. When these constraints are applied to the dataset, the resulting subset is what will become the \textit{components} of the visualization \cite{bertinSemiologyGraphicsDiagrams2011a}. 
%%rework this statement for french labor data
%%In figure~\ref{fig:iris_scatter}, the \textit{invariant} common to all the data being visualized is "sepal length", "petal length", and "species" and the \textit{components} are the measurements of these variables. 
As shown in figure~\ref{fig:flowchart}, the final step in creating a visualization is choosing how to encode the
components using retinal (visual) variables.

\subsubsection{Retinal (Visual) Variables}
\begin{figure}[h!]
\includegraphics[width=1\textwidth]{figures/intro/retinal_variables.png}
\caption{Retinal variables are a codification of how position, size, shape, color and texture are used to illustrate variations in the \textit{components} of a visualization. This tabular form of Bertin's retinal variables is from Understanding Graphics \cite{malamedInformationDisplayTips2010} who reproduced it from \textit{Making Maps: A Visual Guide to Map Design for GIS} 
\cite{krygierMakingMapsVisual2005}}
\label{fig:retinal_variables}
\end{figure}

@@

\subsubsection{Data Type and Structure}
\label{sec:intro_data_structure}
\begin{figure}[h!]
 \includegraphics[width=\textwidth]{figures/intro/munzner_datatypes}
\caption{Keys are unique lookup values used to find individual observations in the dataset. Keys are positional references, and can be coordinates on a map or unique values such as a primary key in a database or a (time, latitude, longitude) index in a data cube. Image modified from a diagram from Munzner's \textit{Visualization Analysis and Design} \cite{munznerVisualizationAnalysisDesign2014}}
\label{fig:munzner_datatypes}
\end{figure}

As shown in figure~\ref{fig:flowchart}, there are multiple ways to translate data into pictures. A map is always an option, except when the observations do not have associated coordinates in a physical plane. Tamara Munzner provides a way to distinguish between these datasets using $\{key, value\}$ designations \cite{munznerVisualizationAnalysisDesign2014}. Munzner defines \textit{values} as measurements of interest in the dataset, analogous to dependent variables in statistics. She defines \textit{keys} as indexes that can be used to look up values, analogous to independent variables in statistics and dimensions in computer science. Figure~\ref{fig:munzner_datatypes} illustrates how these keys are used to look up variables in a dataset: 
\begin{itemize}
	\item map: keys are the coordinates of the points
	\item table: row index, database primary key
	\item data cube: row, column, etc. (.e.g. $i,j,k$) index
\end{itemize}

Expanding on Munzner's key and value semantics, in many datasets the keys are discrete variables like time or geophysical locations sampled from a continuous curve, surface, or field. While these observations are discrete samples from the continuous space, often the continuous (functional) characteristic\cite{ramsayFunctionalDataAnalysis2006a,mullerFunctionalVarianceProcesses2006a} of the observational space is also of interest. Besides quantitative discrete, quantitative continuous, or categorical measurement type considerations, the choice of visualization is also influenced by the measurement being on an interval, ratio, nominal, or categorical scale.

### 2021-01-19 draft/sections/math.tex
removed minted, fleshed out fiber bundles

%% give the why of the math after every mention of why

In this section we introduce a mathematical description of the visualization pipeline where artist $\mathscr{A}$ functions transform data space $\mathscr{E}$ to an intermediate representation in a prerendered graphic space $\mathscr{H}$.

@@

\subsection{Data Space}
\label{sec:data}
Lets say we are working with the Global Historical Climatology Network\cite{AnOverviewoftheGlobalHistoricalClimatologyNetworkDailyDatabase}, which is a global daily record of weather station  such as rainfall and temperature. Depending on which aspects of the data we want to visualize, we can encode it as a timeseries or map or both or even represent the measurements as completely disconnected.

Using a fiber bundle allows us to encode the properties of the heterogenous variables (\ref{sec:data_fiber}), the connectivity of the measurements (\ref{sec:data_base}), and functions between these spaces that yield the measurements (\ref{sec:data_section}).

\subsubsection{Variable Field in F }
\label{sec:data_fiber}
 %%%first sentence should be a paragraph
%%% butler's paper, no notion of distance on K, can only map F to the visual fibers, so need an X to map to Xpos
% variable, measurement, pull in tidy here, 
% % motivate by example, temperature \ time 
%% F_0 = {temp_0, temp_n}, 
%% then adding in a second set of variables time: 
%% F_1 = {time_0, time_1}
% break out into F first as set of values, then F=F_{0}\times F_{1} as component

The fiber is the cartesian product of the variable spaces:
\begin{align}
F &= F_{0} \times \ldots \times F_{n}
\end{align}

We formulate the fiber such that it has a mapping between the somethine thing and the field name, as discussed in Spivak's simplacial description of databases \cite{spivakSIMPLICIALDATABASES,spivakDatabasesAreCategories2010}.

Fibers do not need to be 1D.

\subsubsection{Measurement Type}
%% maybe move somewhere else w/ some more motivation
%% do this section in terms of $M0, F_0$
An important property of data fields are the measurement type, which we generalize to identifying the left monoid action such that the monoid actions $M$ 
\begin{equation}
M = M_{0} \times \ldots\times M_{i}
\end{equation}
is the cartesion cross product of monoids applied to each $F_{i}\in F$. A monoid \cite{Monoid2021} $M$ is a set with an associative binary operator $\ast:M \times M\rightarrow M$. A monoid has an identity element $e\in M$ such that $e\ast a= a \ast e = a$ for all $a \in M$. A left monoid action \cite{SemigroupAction2021,ActionNLab} of $M$ is a set $F$ with an action $\bullet: M\times F \rightarrow F$ with the properties:
\begin{description}
    \item[associativity] $\text{for all } m,t \in M \text{ and } x\in F, m\bullet(t\bullet x) = (m\bullet t) \bullet x$
    \item[identity] $\text{for all } x\in F, e\in M,  e\bullet x = x$ 
\end{description}

@@

Datasets have a semantic topology such as the values are interpreted as discrete observations or part of a timeseries or map, or nodes in a networks \cite{munznerWhatDataAbstraction2014,geveci2012vtk}. %%turn it into a paragraph

As illustrated in fig~\ref{fig:base_space_div}, $K$ is this underlying structure. $K$ does not directly know the values; instead it is the sections that define the lookup between keys $k \in K$ and the corresponding values in $E$.

@@

\paragraph{Q: scatter plot}
\begin{figure}[h!]
    \includegraphics[width=\textwidth]{figures/math/scatter.png}
    \label{fig:artist_scatter}
\end{figure}

@@

\paragraph{Q: line plot}
\begin{figure}[h!]
    \includegraphics[width=\textwidth]{figures/math/line.png}
    \label{fig:artist_line}
\end{figure}

The line plot shown in fig~\ref{fig:artist_scatter} exemplifies the need for the jet discussed in section~\ref{sec:sheaf_stalk}

@@

which yields components of $\rho(s)$:
\begin{align}
 x = xpos(\xi(\alpha_1)) &+ \alpha_2\hat(n_1)(\xi(\alpha_1)) \\
 y = ypos(\xi(\alpha_1)) &+ \alpha_2\hat(n_2)(\xi(\alpha_2)) 
\end{align}

\paragraph{Q: heatmap}
\begin{figure}[h!]
    \includegraphics[width=\textwidth]{figures/math/heatmap.png}
    \label{fig:artist_heatmap}
\end{figure}
The heatmap in figure~\ref{fig:artist_heatmap} 
\begin{equation}
Q(xpos, ypos, color)
\end{equation}
has in the simple case a direct lookup into $K$ to obtain the $\mu = (x,y,c)$ values that are mapped into 
\begin{align}
D_{RGB} &= color(\xi(\alpha_1, \alpha_2))
D_x & = xpos(\xi(\alpha_1, \alpha_2))
D_y &= ypos(\xi(\alpha_1, \alpha_2))
\end{align}
through $\rho$.

@@

One way to build flexible $\xi$ is to choose a consistent way of representing$K$. In our draft implementation of the data as fiber bundle model, we triangularize $K$ using complexes of the simplicies shown in figure~\ref{fig:triangle_simplex} such that $\xi$ consistently yields some combination of vertexes, edges, and faces. This gives a common data indexing structure on which to build components that could potentially be reused across $Q$.

>>>>

%% give the why of the math after every mention of why 
In this section we introduce a mathematical description of the visualization pipeline where artist $\mathscr{A}$ functions transform data space $\mathscr{E}$ to an intermediate representation in a prerendered graphic space $\mathscr{H}$.

@@

\subsection{Data Space}
\label{sec:data}
Lets say we are working with the Global Historical Climatology Network\cite{AnOverviewoftheGlobalHistoricalClimatologyNetworkDailyDatabase}, which is a dataset of global daily meterological observations. Each record includes the timestamp, location, and measurements such as temperature, precipitation, and snowfall. Using a fiber bundle allows us to express the properties of the variables(\ref{sec:data_fiber}), the connectivity of the observations (\ref{sec:data_base}), and tables of records (\ref{sec:data_section}).

\subsubsection{Variable Field in F }
\label{sec:data_fiber}
The fiber is a topological space containing the set of all observations that could possible be in the dataset, even values that do not actually exist in the dataset. For example, lets construct a fiber bundle that is just the temperature column from GHCN. Then the fiber $F$ is the range of physically possible temperature values $\text{Temp} \subset \mathbb{R}$. Lets rename this
\begin{equation}
F_0 = \{\text{Temp}\}
\end{equation}
so that more variables can be added to the fiber. For example, lets add the time from GHCN $F_1 =\text{Time}$
\begin{equation}
F =  \{\text{Temp}\} \times \text{Time}
\end{equation}

The elements in a given fiber set do not need to be scalers; for example the location can be encoded as $F_3=\text{Loc}$ where $(\text{lat},\text{lon{}) \in \text{Loc}$.  We add the location to the fiber by adding the set to the cartesian product
\begin{equation}
    F =  \{\text{Temp}\} \times \text{Time} \times \text{Loc}
\end{equation}

We also add the variable name and measurement scale $M$ monoid to the fiber $F_i$ such that 
\begin{align}
    $F_0$ &= (\mathrm{temperature}, \text{Temp}, $M_0$)\\
    $F_1$ &= (\mathrm{time}, \text{Time}, $M_1$)\\
    $F_2$ &= (\mathrm{location}, \text{Loc}, $M_2$)
\end{align}

as described in Spivak's categorical formulation of databases \cite{spivakSIMPLICIALDATABASES}. We describe $M$ in more detail in section~\ref{sec:data_monoid}. We can keep growing the fiber
\begin{equation}
F= F_{0} \times \ldots \times F_{n}
\end{equation}
while also retaining the individual sets $F_{i}$ that we will later apply visual variables to.

\subsubsection{Measurement Type}
\label{sec:data_monoid}
%% maybe move somewhere else w/ some more motivation
%% do this section in terms of $M0, F_0$
An important property of data fields are the measurement type, which we generalize to identifying the left monoid action such that the monoid actions $M$ 
\begin{equation}
M = M_{0} \times \ldots\times M_{i}
\end{equation}
is the cartesion cross product of monoids applied to each $F_{i}\in F$. A monoid \cite{Monoid2021} $M$ is a set with an associative binary operator $\ast:M \times M\rightarrow M$. A monoid has an identity element $e\in M$ such that $e\ast a= a \ast e = a$ for all $a \in M$. A left monoid action \cite{SemigroupAction2021,ActionNLab} of $M$ is a set $F$ with an action $\bullet: M\times F \rightarrow F$ with the properties:
\begin{description}
    \item[associativity] $\text{for all } m,t \in M \text{ and } x\in F, m\bullet(t\bullet x) = (m\bullet t) \bullet x$
    \item[identity] $\text{for all } x\in F, e\in M,  e\bullet x = x$ 
\end{description}

@@

As mentioned in section~\ref{sec:intro_data_structure}, datasets have a semantic topology wherein the values are interpreted as discrete observations or part of a timeseries or map, or nodes in a networks \cite{munznerWhatDataAbstraction2014,geveci2012vtk}. one way of encoding this underlying structure are the $K$s illustrated in figure~\ref{fig:base_space_div}, $K$ does not directly know the values; instead it is the sections that define the lookup between keys $k \in K$ and the corresponding values in $E$.

@@

\paragraph{Q: scatter plot}
\\
\begin{figure}[h!]
    \includegraphics[width=\textwidth]{figures/math/scatter.png}
    \label{fig:artist_scatter}
\end{figure}

@@

\paragraph{Q: line plot}
\\
\begin{figure}[h!]
    \includegraphics[width=\textwidth]{figures/math/line.png}
    \label{fig:artist_line}
\end{figure}

The line plot shown in fig~\ref{fig:artist_scatter} exemplifies the need for the jet discussed in section~\ref{sec:sheaf_stalk}
(needs the normal to push up/down against the normal)

@@

%% only alpha1 exists on the data side
which yields components of $\rho(s)$:
\begin{align}
 x = xpos(\xi(\alpha_1)) &+ \alpha_2\hat(n_1)(\xi(\alpha_1)) \\
 y = ypos(\xi(\alpha_1)) &+ \alpha_2\hat(n_2)(\xi(\alpha_2)) 
\end{align}

\paragraph{Q: heatmap}
\\
\begin{figure}[h!]
    \includegraphics[width=\textwidth]{figures/math/heatmap.png}
    \label{fig:artist_heatmap}
\end{figure}
The heatmap in figure~\ref{fig:artist_heatmap} 
\begin{equation}
Q(xpos, ypos, color)
\end{equation}
has in the simple case a direct lookup into $K$ to obtain the $\mu = (x,y,c)$ values that are mapped into 
(arrays are upside down, x is y, is x) - $\xi$ is about translating indices from data to visual alignment
\begin{align}
D_{RGB} &= color(\xi(\alpha_1, \alpha_2))
D_x & = xpos(\xi(\alpha_1, \alpha_2))
D_y &= ypos(\xi(\alpha_1, \alpha_2))
\end{align}
through $\rho$.

@@

One way to build flexible $\xi$ is to choose a consistent way of representing $K$. In our draft implementation of the data as fiber bundle model, we triangularize $K$ using complexes of the simplicies shown in figure~\ref{fig:triangle_simplex} such that $\xi$ consistently yields some combination of vertexes, edges, and faces. This gives a common data indexing structure on which to build components that could potentially be reused across $Q$.
